---
title: "Model Types and Diagnostics"
date: April 30, 2025
execute:
  warning: false
  message: false
  fig-align: 'center'
  out-width: 75%
format:
  html:
    html-math-method: katex
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: true
    code-summary: "Show the code"
---

```{r packages}
#| echo: false
#| code-fold: false
library(tidyverse)
library(regressinator)
library(broom)
library(janitor)
library(DHARMa)
library(sandwich)
library(rsample)
library(car)
library(lmtest)
library(lme4)
library(mgcv)
library(splines)
library(glmnet)
library(gt)
library(gtsummary)
library(kableExtra)
library(sf)
library(GGally)
library(patchwork)
library(ggeffects)
library(ggthemes)
```

# Linear Regression

## Assumptions

For the linear regression model $\mathbf{Y} = \mathbf{X}\beta + e$, we have the following assumptions: \

(1) Errors have mean 0: $\mathbb{E}[Y \mid X] = 0$ \
(2) The error variance is constant: $\text{Var}[e \mid \mathbf{X}] = \sigma^2$ \
(3) The errors are uncorrelated (the data points are iid): $\text{Var}[e \mid \mathbf{X}] = \sigma^2$\
(4) The errors are normally distributed


## Diagnostics and Solutions 

### Mean Model Misspecification

**Linearity:** partial residual plots help us diagnose whether the true relationship between the outcome and some of the predictors is non-linear

```{r}
#| echo: false

rail_trails <- read_csv("data/rail-trails.csv") |>
  rename(housenum = `housenum...1`) |>
  select(-`housenum...15`) |>
  mutate(across(where(is.character), as.factor),
         zip = as.factor(zip))

# dropping the columns that are descriptive/metadata but not useful for modeling
rail_trail_modeling <- rail_trails |>
  select(-housenum, -longitude, -latitude, -pctchange, 
         -streetno, -contains("2014"), -contains("2011"),
         -contains("1998"), -adj2007) 

df_residuals <- lm(walkscore~distance+I(distance^2), data = rail_trail_modeling) |> augment() 

rail_trail_modeling <- rail_trail_modeling |>
  cbind(resid_walkscore = df_residuals$.resid) 
```

```{r}
#| label: fig-partial-resids
fit <- lm(price2007 ~ distance + resid_walkscore + squarefeet + bedgroup + 
            zip, data = rail_trail_modeling)

fit_new <- lm(price2007 ~ distance + ns(resid_walkscore, knots = c(0.5)) + 
                squarefeet + bedgroup + zip, data = rail_trail_modeling)

p1 <- partial_residuals(fit) |>
  mutate(.predictor_name = case_when(
    .predictor_name == "distance" ~ "Distance to Rail Trail",
    .predictor_name == "walkscore" ~ "Walking Score",
    .predictor_name == "bikescore" ~ "Biking Score",
    .predictor_name == "squarefeet" ~ "Square Footage",
    .predictor_name == "resid_walkscore" ~ "Residual Walk Score",
    TRUE ~ .predictor_name)) |>
  ggplot(aes(x = .predictor_value, y = .partial_resid)) +
  geom_point(color = "grey40", size = 0.5) +
  geom_smooth(color = "blue", fill = 'skyblue2',
              linewidth = 0.7, linetype = "twodash") +
  geom_line(aes(y = .predictor_effect), color = 'red') +
  facet_wrap(vars(.predictor_name), scales = "free", ncol = 2) +
  labs(x = "Predictor value", y = "Partial residual")+
  ggtitle("(a) Partial Residual Plots,\nPreliminary Linear Model")+
  theme_bw()+
  theme(strip.background = element_rect(fill = "white", color = "white"),
        strip.text = element_text(color = "black", face = "bold"),
        axis.text = element_text(color = "black"),
        plot.title = element_text(face = "bold", size = 11))

p2 <- partial_residuals(fit_new) |>
  mutate(.predictor_name = case_when(
    .predictor_name == "distance" ~ "Distance to Rail Trail",
    .predictor_name == "walkscore" ~ "Walking Score",
    .predictor_name == "bikescore" ~ "Biking Score",
    .predictor_name == "squarefeet" ~ "Square Footage",
    .predictor_name == "resid_walkscore" ~ "Residual Walk Score",
    TRUE ~ .predictor_name)) |>
  ggplot(aes(x = .predictor_value, y = .partial_resid)) +
  geom_point(color = "grey40", size = 0.5) +
  geom_smooth(color = "blue", fill = 'skyblue2',
              linewidth = 0.7, linetype = "twodash") +
  geom_line(aes(y = .predictor_effect), color = 'red') +
  facet_wrap(vars(.predictor_name), scales = "free", ncol = 2) +
  labs(x = "Predictor value", y = "Partial residual")+
  ggtitle("(b) Partial Residual Plots, Adding\nNatural Splines (Knot at Median)\nto Residual Walk Score")+
  theme_bw()+
  theme(strip.background = element_rect(fill = "white", color = "white"),
        strip.text = element_text(color = "black", face = "bold"),
        axis.text = element_text(color = "black"),
        plot.title = element_text(face = "bold", size = 11))

p1 + p2
```


From @fig-partial-resids(a), we see that residual walk score has a nonlinear relationship with estimated house price in 2007 -- as indicated by the smoothed residual curve (in blue) diverging from the predicted effect line (in red). We would address this nonlinearity by fitting natural splines on the residual walk score, with one knot at the median. @fig-partial-resids(b) provides the updated partial residual plots when we add natural splines to the residual walk score. We see now that the predicted effect line (in red) tracks closely to the smoothed residual curve (in blue) without over-fitting to model to the turns in the smoothed residual curve. 

**Lineups:** If we aren't sure whether the true relationship between the outcome and a given predictor is non-linear, we can conduct a partial residual visual lineup.

For instance, it is not entirely clear whether distance has a non-linear relationship to estimated house price in 2007 or if this is just due to the high-leverage home with the largest distance to the rail trail. @fig-lineup-distance shows the partial residual line-up. If we cannot pick out the partial residual plot corresponding to the real data from the simulated data, we can conclude that there is  insufficient evidence of a non-linear relationship between distance to rail trail and estimated house price in 2007 at a significance level of 0.05.

```{r}
#| message: false
#| label: fig-lineup-distance
set.seed(123)
model_lineup(fit_new, fn = partial_residuals, nsim = 20) |>
  filter(.predictor_name == "distance") |>
  mutate(.sample_new = paste0("Position ", .sample),
         .sample_new = fct_reorder(.sample_new, .sample)) |>
  ggplot(aes(x = .predictor_value, y = .partial_resid))+
  geom_point(color = "grey40", size = 0.5) +
  geom_line(aes(y = .predictor_effect), color = 'red') +
  geom_smooth(color = "blue", fill = 'skyblue2',
              linewidth = 0.7, linetype = "twodash") +
  facet_wrap(~.sample_new)+
  labs(x = "Distance to Rail Trail (in Miles)", y = "Partial residual")+
  theme_bw()+
  theme(strip.background = element_rect(fill = "white", color = "white"),
        strip.text = element_text(color = "black", face = "bold"),
        axis.text = element_text(color = "black"),
        plot.title = element_text(face = "bold", size = 11))
```

We can decrypt (show the the position of our actual data) by running the command, which is given at the end of the line-up call (`dycrypt()`):

```{r}
#| echo: true
#| code-fold: false
decrypt("QUg2 qFyF Rx 8tLRyRtx ZP")
```


#### Dealing with Nonlinear Predictors

1. Add polynomial terms to the model

Simplest way to add polynomial terms is with `I()`:

```{r}
#| echo: true
#| code-fold: false

fit_new <- lm(price2007 ~ distance + resid_walkscore + I(resid_walkscore^2) +
                squarefeet + bedgroup + zip, data = rail_trail_modeling)
```

If there is numerical issues (e.g., from the predictor having large values), we can use the `poly()`function which constructs the design matric using an orthogonal polynomial basis. 

**A downside is that the coefficients are no longer is terms of the interpretable regressor terms. The fitted values though wil be the same between the two approaches for adding polynomial terms**. 

```{r}
#| echo: true
#| code-fold: false
fit_new <- lm(price2007 ~ distance + poly(resid_walkscore, degree = 2) + squarefeet + 
                bedgroup + zip, data = rail_trail_modeling)
```

2. Add regression splines to the model

Regression splines model relationships as being piecewise polynomial. They require us to choose knots, which are the fixed points between which the function is polynomial. Two common approaches are *natural splines* and *B-spline basis*. Statisticians almost universally use cubic splines. Cubic splines are continuous in their first and second derivatives, making the knots almost visually imperceptible in plots of the spline.

For knot selection, a typical approach is to use quantiles of the data as knots -- erring on the side of less knots (especially if there is few observations).

```{r}
#| echo: true
#| code-fold: false

fit_new <- lm(price2007 ~ distance + ns(resid_walkscore, knots = c(0.5)) + 
                squarefeet + bedgroup + zip, data = rail_trail_modeling)
```


### Constant Error Variance

Errors with non-constant variance are called *heteroskedastic*. To detect heteroskedasticity, we can examine the residuals plotted against the fitted values. 

```{r}
#| label: fig-constant-errors
fit_new |>
  augment() |>
  ggplot(aes(x = .fitted, y = .resid))+
  geom_point(size = 0.5)+
  geom_hline(yintercept = 0)+
  geom_smooth(fill = 'skyblue2')+
  labs(x = "Fitted Values", y = "Residuals")+
  theme_bw()+
  theme(axis.text = element_text(color = "black"))
```

In @fig-constant-errors, we have non-constant error variance. Indeed, there appears to be an approximately quadratic relationship between our fitted values and our residual terms. 

**We can remedy non-constant error variance by using the sandwich estimator for the variance of our estimators in all inference.**

```{r}
#| echo: true
#| code-fold: false
# uses sandwich estimator with HC3 estimator for Omega
Confint(fit_new, vcov = vcovHC(fit_new))
```

### Normality of Errors

We can check for non-normality of errors using residual Q-Q plots.  Using the standardized residuals, we can detect gross deviations from normality while ignoring small deviations in non-normality. An example of a Q-Q plot for linear regression is in @fig-qq-plot. Since we are doing linear regression, the theoretical quantiles correspond to a *normal distribution with mean 0 and variance 1*.

```{r}
#| label: fig-qq-plot
augment(fit_new) |>
  ggplot(aes(sample = .std.resid)) +
  geom_qq_line(linewidth = 0.4) +
  geom_qq(size = 0.75)+
  labs(x = 'Theoretical Quantiles',
       y = 'Observed Quantiles')+
  theme_bw()
```


Like with partial residuals, we can use a lineup to detect whether the non-normality in errors is statistically significant (at a significance level of 0.05). 

```{r}
set.seed(123)
model_lineup(fit_new) |> 
  mutate(.sample_new = paste0("Position ", .sample),
         .sample_new = fct_reorder(.sample_new, .sample)) |> 
  ggplot(aes(sample = .std.resid)) +
  geom_qq_line(linewidth = 0.4) +
  geom_qq(size = 0.5) +
  facet_wrap(vars(.sample_new)) +
  labs(x = "Theoretical Quantiles",
       y = "Observed Quantiles")+
  theme_bw()+
  theme(strip.text = element_text(face = "bold"),
        strip.background = element_rect(fill = "white", color = "white"),
        text = element_text(color = "black"),
        axis.text = element_text(color = "black"))
```

```{r}
#| echo: false
#| code-fold: false
decrypt("QUg2 qFyF Rx 8tLRyRtx ZP")
```


*Note:* if the sample size is sufficiently large, then violating non-normality of errors is not a serious concern. Otherwise, if the assumptions linearity and constant error variance hold, then interpret coefficient estimates with caution (since coefficient estimates would not be approximately normal).  


## Contaminated Errors and Outliers

**Outliers:** Not all outliers have a major effect on the regression line. So, their presence may not be a problem. We hence need a way to detect outliers and characterize their influence on the regression line.

  (1) If we can determine the outlier is due to some kind of measurement or recording error, we can correct the error.
  
  (2) If we cannot, we must make the difficult decision of whether to keep the outlier, and acknowledge that it may significantly influence our estimates, or to remove it and risk throwing away good information
  

One way to measure the influence of an observation on the regression is to quantify how its inclusion changes our coefficient estimates via *Cook's Distance*.

- Look for Cook's Distance ($D_i$), where $D_i \geq 1$, though this is again a matter of judgment, to indicate that a particular observation substantially changes the regression fit.

@fig-cooks-distance shows that all observations have Cook's Distances well below the standard cut-off of $1$, so there does not appear that any of the homes in the data have a large influence on our regression model.  

```{r}
#| label: fig-cooks-distance
max_price <- augment(fit_new) |>
  arrange(desc(price2007)) |>
  slice_head(n = 1)

augment(fit_new) |>
  ggplot(aes(y = .cooksd, x = price2007))+
  geom_point(size = 0.7, alpha = 0.7)+
  geom_point(data = max_price, aes(y = .cooksd, x = price2007),
             size = 2, color = "red")+
  labs(y = "Cook's Distance", x = "Estimated House Price (2007)")+
  #geom_hline(yintercept = 1, linetype = "dashed")+
  theme_bw()+
  theme(plot.title = element_blank(),
        text = element_text(color = "black"),
        axis.text = element_text(color = "black"),
        axis.title = element_text(color = "black"))
```

## Collinearity

When the model matrix $X$ is not full column rank, its columns span a lower-dimensional space. There is still a unique perpendicular projection of $Y$ onto this space, and hence a unique $\hat Y$. *But* the coordinates of that fit are not unique, meaning that there are infinite values of $\hat \beta$ that correspond to the same squared error/prediction. **If we care about inference on $\hat \beta$ this is a serious problem**. 

```{r}
loess_wrapper <- function(data, mapping, ...){
      p <- ggplot(data = data, mapping = mapping) + 
      geom_point(size = 0.35, color = "skyblue2") + 
      geom_smooth(linetype = "dashed", color = "black", linewidth = 0.6)
      return(p)
}

rail_trail_modeling |>
  select(price2007, distance, walkscore, squarefeet) |>
  rename(`2007 Est.\nHouse Price\n(thousands of $)` = price2007,
         `Distance to\nRail Trail (miles)` = distance,
         `Walk\nScore` = walkscore,
         `Square\nFootage` = squarefeet) |>
  ggpairs(upper = list(continuous = wrap("cor", color = "black")),
          lower = list(continuous = loess_wrapper,
                       combo = wrap("barDiag", size=0.2,
                                     alpha = 0.9))) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 4))+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 4))+
  theme_bw()+
  theme(strip.background = element_rect(fill = "white", color = "white"),
        strip.text = element_text(color = "black", face = "bold", size = 6.5),
        axis.text = element_text(color = "black", size = 6.5))
```


### Non-pairwise Collinearity Metrics

**Condition Number:**

However, multiple variables can be collinear even when individual pairs of variables have low correlation. One way to detect non-pairwise collinearity is to use the condition number, where we compare the largest and smallest eigenvalues of $X^\intercal X$:

```{r}
#| echo: true
#| code-fold: false

X <- model.matrix(price2007 ~ distance + resid_walkscore + 
                squarefeet + bedgroup + zip - 1, data = rail_trail_modeling)
# first step is to scale the model matrix X
scaled_x <- scale(X, center = FALSE, scale = TRUE)
  
# Then, we can use the kappa() function on X'X
kappa(t(scaled_x) %*% scaled_x)
```

Ideally, we want the condition number to be small. The general cut-off is condition numbers larger than $50$ or $100$. So, from the results above, we see that there is potentially concerning levels of collinearity among our predictors, which could influence the accuracy of our inference.

**Variance Inflation Factors:**

Another way to look at collinearity is with variance inflation factors (VIF), which can be interpreted as showing how much the variance of $\hat \beta_j$ is inflated relative to a model where there is no collinearity (and all columns are orthogonal). 
```{r}
#| echo: true
#| code-fold: false
# consider using type = 'predictor' for interactions (or ignoring VIFs since compared to main effecs)
car::vif(lm(price2007 ~ distance + ns(resid_walkscore, knots = c(0.5)) + 
                squarefeet + bedgroup + zip, data = rail_trail_modeling))
```

A general rule of thumb is that VIF values exceeding 5 warrant further investigation, and VIFs exceeding 10 are signs of serious multicollinearity requiring correction.

**General Notes:** If we are confronted with collinearity, then, the question to ask is:

  - "Have we chosen the correct predictors for the research question?"
  
If we have, there is little to be done; if we have not, we can reconsider our choice of predictors and perhaps eliminate the collinear ones.

If we are interested in prediction rather than in the coefficients, *collinearity is a problem insofar as it creates high prediction variance, and we might reconsider our model and use a penalization model to reduce the prediction variance.*s

## Interactions

An interaction allows one predictor’s association with the outcome to depend on values of another predictor.

When an interaction is present, the normal interpretation of coefficients as slopes no longer holds for the predictors involved in the interaction.

**See Example 7.4**

## Hypothesis Tests

### t Tests

Many useful null hypotheses can be written in terms of linear combinations of the coefficients. For example, consider a linear model with a continuous predictor $X_1$ and dummy-coded regressor $X_2 \in \{0,1 \}$:

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2$$
We could think of several null hypotheses to test:

- $\beta_2 = 0$: the two factor levels have identical intercepts
- $\beta_3 = 0$: the two factor levels have identical slopes
- $\beta_1 = c$: the slope when $X_2 = 0$ is some value $c$ predicted by a theory we are testing (usually $c = 0$ in linear regression, which corresponds to no association)
- $\beta_1 + \beta_3 = 0$: when $X_2 = 1$, there is no association between $X_1$ and $Y$
- $\beta_2 = \beta_3 = 0$: the two factor levels have identical relationships between $X_1$ and $Y$

For all but the last listed hypothesis, we can use a t test. `R` conducts a t test for every coefficient by default, with $c = 0$. 

The **degrees of freedom for a t test is $n - p$, where $p$ is the number of parameters in our model.** We can get this via `fit$df.residual`. E.g., for our rail trail linear model example we have $df = 96$.

The function `tidy()` does the test for $c = 0$ with every other predictor held constant in the model.

```{r}
gt(tidy(fit_new)) |>
  fmt_number(decimals = 3)
```


We can also use `tbl_regression()` which gives us the confidence intervals and p-values:

```{r}
tbl_regression(fit_new)
```

**If we have non-constant error variance and thus use the sandwich estimator, we can use the following code to conduct t tests and present our results:**

```{r}
#| echo: true
#| code-fold: false
p.values <- coeftest(fit_new, vcov = vcovHC(fit_new)) |>
  tidy() |> _$p.value

t.stats <- coeftest(fit_new, vcov = vcovHC(fit_new)) |>
  tidy() |> _$statistic

data_table <- as.data.frame(Confint(fit_new, vcov = vcovHC(fit_new))) |>
  cbind(t.stats) |>
  cbind(p.values) |>
  rownames_to_column(var = "Characteristic") |>
  rename(`p-value` = p.values,
         `t-stat` = t.stats)

gt(data_table) |>
  fmt_number(decimals = 3)
```

Of course, a fourth option is to manually perform the t test, which might make sense if we have a complex null or would like to do a one-sided test. Below is the general code for this:

```{r}
#| eval: false
#| echo: true
#| code-fold: false

beta_hats <- coef(gentoo_fit)
a <- c(0, 0, 0, 1) # beta_3 = 0

t_stat <- sum(a * beta_hats) /
  sqrt(a %*% vcov(gentoo_fit) %*% a)[1,1]

p_value <- pt(t_stat, gentoo_fit$df.residual,
              lower.tail = FALSE)

p_value * 2 # if two-sided
```


### F tests for Nested Models

The `anova()` function can be used to conduct F tests for nested models. For instance, we can test whether there was a significant association between residual walk score and estimated home price in 2007 (in practice, probably not a good idea since residual walk score is random):

```{r}
# create a model without the predictor
fit_no_resid_ws <- lm(price2007 ~ distance + 
                squarefeet + bedgroup + zip, data = rail_trail_modeling)

anova(fit_no_resid_ws, fit_new)
```

**The parameters of an F-statistic are $q$ and $n - p$. `Df` is the outputted table corresponds to $q$ and the second row of `Res.Df` corresponds to `n - p`**. E.g, we have $F(2, 96)$ as our null distribution in this example.

## Coefficient Interpretation

**In a report, we just include full interpretations for the hypothesis tests and/or coefficient estimates that relevant to the practical problem at hand!**

```{r}
#| label: tbl-penguin-results
#| tbl-cap: ADD
library(palmerpenguins)

penguin_fit <- lm(
  bill_length_mm ~ flipper_length_mm + species +
    flipper_length_mm:species,
  data = penguins
)
tbl_regression(penguin_fit)
```


### Intercept

The intercept $\beta_0$ is the mean value of  when all other regressors are 0:

$$\mathbb{E}[Y \mid X_1 = 0, \dots, X_q = 0]$$
Interpreting the intercept, quoting confidence intervals, and conducting hypothesis tests for its value may only be useful when the intercept has a substantive meaning and we have observed $X$ values nearby.


### Slopes

*Note these are the general parts of an interpretation for a continuous predictor in a linear regression:*

- Difference in the mean value of $Y$

- Associated with a one-unit increase in $X_1$

- Holding all other regressors constant

**Exact interpretation depends on if the predictor is continuous or a factor level:**

- **Test for a factor coefficient:** For a given flipper length, gentoo penguins have a smaller mean bill length ($M = 47.5$ mm, $SD = 3.1$) than Adelie penguins ($M = 38.8$ mm, $SD = 2.7$), $t(336) = -3.5, p = 0.001$.

*In part calculated using `group_by()`:*

```{r}
penguins |>
  group_by(species)|>
  summarize(mean_bill_length =
              mean(bill_length_mm,
                 na.rm = TRUE)) |>
  gt() |>
  fmt_number(decimals = 1)
```


- **Confidence interval for a factor coefficient:** For a given flipper length, gentoo penguin bills are shorter than Adelie penguin bills by an average of $34.3$ mm (95% CI $[15.01, 15.64]$). 

- **Test for slope:** In Adelie penguins, there was a statistically significant association between flipper length and bill length, $\hat \beta = 0.13$, $t(336) = 4.17, p < 0.001$.

- **Describe a slope:** Among Adelie penguins, each additional millimeter of flipper length is associated with $0.13$ mm of additional bill length, on average (95% CI $[0.07, 0.2]$).

### Interactions

**Describing an interaction coefficient:** For the interaction term between flipper length and species from @tbl-penguin-results: For Adelie penguins (the baseline level), the association between bill length and flipper length is $0.13$ mm of bill length per millimeter of flipper length, on average (95% CI $[0.07, 0.20]$). But for Chinstrap penguins, the association is $0.13 + 0.09 = 0.22$ mm of bill length per millimeter of flipper length, on average (95% CI $[0.11, 0.31]$). [**LOOK AT THEOREM 5.5 AND HW 3 TO CONFIRM**]

**A helpful way to visualize interaction terms, polynomial terms, (or other complex relationships) is to use an effects plot:**

```{r}
#| echo: true
#| code-fold: false
predict_response(penguin_fit,
                 terms = c("flipper_length_mm", "species")) |>
  plot() +
  labs(x = "Flipper length (mm)", y = "Bill length (mm)",
       color = "Species",
       title = "Flipper length effect plot")
```

*In an effects plot, non-focal predictors are set to their mean (numeric variables), reference level (factors), or "most common" value (mode) in case of character vectors.*


### Spline or Polynomial Terms

If we fit a spline or a polynomial for our predictor-of-interest, we cannot determine the statistical significance or effect size from the outputted regression tables. Instead, we must use the following framework:

1. Do a LRT/F-test for nested models to obtain the overall significance of the predictor-of-interest.

```{r}
#| echo: true
#| code-fold: false
anova(fit_new, fit_no_resid_ws)
```


2. Create an effects plot fort the variable of interest 


```{r}
#| echo: true
#| code-fold: false
predict_response(fit_new,
                 terms = c("resid_walkscore")) |>
  plot()
```


3. Give an example of an interpretation using the effects plot with specific values


```{r}
#| echo: true
#| code-fold: false
preds <- predict_response(fit_new, terms = c("resid_walkscore"))
head(preds)
```

For example, when the residual walk score is $-43.01$, the predicted house price (in 2007) is \$259,570 in Northampton, MA (95% CI $[\$213,440, \$305,710]$) versus when the residual walk score is $-36.08$, the predicted house price (in 2007) is \$271,780 in Northampton, MA (95% CI $[\$233,600, \$309,960]$) -- holding all other covariates at their mean or reference level.

# Logistic Regression

When $Y \in \{0, 1\}$, $X \in \mathbb{R}^p$, and $\beta \in \mathbb{R}^p$, logistic regression models the response as:

$$\text{logit(Pr}(Y = 1 \mid X = x)) = \beta^\intercal x$$
which is equivalent to saying:

$$
\begin{aligned}
\text{Pr}(Y = 1 \mid X = x) &= \text{logit}^{-1}(\beta^\intercal x)\\
\text{odds}(Y = 1 \mid X = x) &= \exp(\beta^\intercal x)\\
\log(\text{odds}(Y = 1 \mid X = x)) &= \beta^\intercal x
\end{aligned}
$$

and notably:

$$\hat y_i = \widehat{\text{Pr}}(Y = 1 \mid X = x_i) = \text{logit}^{-1}(\hat \beta^\intercal x_i)$$

## Assumptions

Logistic regression makes two basic assumptions about the population relationship:

(1) The log-odds is linearly related to the regressors: $\log(\text{odds}(Y = 1 \mid X = x)) = \beta^\intercal x$

(2) The observation $Y_i$ are conditionally independent given the covariates $X_i$


## Linearity Diagnostics

**Empirical Link Plot (EDA):**

**Example:** @fig-logistic-empirical-link shows that TWA core temperature during surgery appears to have an approximately linear relationship with the log-odds of superficial post-op infection but possibly a non-linear relationship with log-odds of a serious post-op infection.


```{r}
#| label: fig-logistic-empirical-link
core_temperature <- read_csv("data/core-temperature.csv") |>
  select(-SurgeryType) |>
  filter(DEAD == 0 | DurationHosp >= 30)

p1 <- core_temperature |>
  bin_by_quantile(TWATemp, breaks = 10) |>
  summarize(
    mean_temp = mean(TWATemp),
    prob = mean(SeriousInfection),
    log_odds = empirical_link(
      SeriousInfection,
      family = binomial(link = "logit")),
    `Patient Count` = n()
  ) |>
  ggplot(aes(x = mean_temp, y = log_odds)) +
  geom_point(aes(size = `Patient Count`))+
  geom_smooth(fill = "skyblue3")+
  scale_size_continuous(range = c(1,3))+
  labs(title = "(A) Serious Infection",
       y = "Log-odds of Infection",
       x = "TWA Core Temperature during\nSurgery (in Celsius)")+
  theme_minimal()+
  theme(legend.position = "None",
        plot.title = element_text(face = "bold", size = 11, hjust = 0),
        axis.text = element_text(color = "black"))

p2 <- core_temperature |>
  bin_by_quantile(TWATemp, breaks = 10) |>
  summarize(
    mean_temp = mean(TWATemp),
    prob = mean(SuperficialInfection),
    log_odds = empirical_link(
      SuperficialInfection,
      family = binomial(link = "logit")),
    `Patient Count` = n()
  ) |>
  ggplot(aes(x = mean_temp, y = log_odds)) +
  geom_point(aes(size = `Patient Count`))+
  geom_smooth(fill = "skyblue3")+
  scale_size_continuous(range = c(1,3))+
  labs(title = "(B) Superficial Infection",
       y = "Log-odds of Infection",
      x = "TWA Core Temperature during\nSurgery (in Celsius)")+
  theme_minimal()+
  theme(plot.title = element_text(face = "bold", size = 11, hjust = 0),
        axis.text = element_text(color = "black"))

p1 + p2
```

**Partial Residuals:**

Another method is to use partial residuals, just like we did for linear regression.  

**Example:** @fig-prelim-partials shows that surgery duration and TWA core temperature both appear to have non-linear relationships with the log-odds of serious infection, which confirms what we see in the empirical link plot.


```{r}
#| label: fig-prelim-partials
core_temperature_model <- core_temperature |>
  filter(DEAD == 0 | DurationHosp >= 30) |>
  mutate(WeightLoss = ifelse(WGHTLOSS == 1, "Yes", "No"),
         SteroidUsage = ifelse(SteroidHx == 1, "Yes", "No"),
         SurgeryDuration = SurgDuration)

prelim_fit <- glm(SeriousInfection ~ TWATemp + WeightLoss +
                    SurgeryDuration + SteroidUsage,
    data = core_temperature_model,
    family = binomial()) 

partial_residuals(prelim_fit) |>
    mutate(.predictor_name = case_when(
    .predictor_name == "Age" ~ "Age (in Years)",
    .predictor_name == "BMI" ~ "Body Mass Index (kg/m^2)",
    .predictor_name == "SurgeryDuration" ~ "Surgery Duration (in Minutes)",
    .predictor_name == "TWATemp" ~ "TWA Core Temperature (in Celsius)",
    TRUE ~ .predictor_name
  )) |>
  ggplot((aes(x = .predictor_value, y = .partial_resid)))+
  geom_point(size = 0.3, color = 'grey50')+
  geom_line(aes(y = .predictor_effect), linewidth = 0.3)+
  geom_smooth(method = "loess", linetype = "twodash", fill = "skyblue3")+
  facet_wrap(~.predictor_name, scales = "free")+
  labs(y = "Partial Residual",
       x = "Predictor Value")+
  theme_minimal()+
  theme(strip.text = element_text(face = "bold"))
```

We address the non-linearity in @fig-prelim-partials by fitting natural splines to both surgical duration and TWA core temperature. 

```{r}
#| label: fig-updated-partials

updated_fit <- glm(SeriousInfection ~ splines::ns(TWATemp, knots = c(35.7, 36, 36.4)) + WeightLoss + ns(SurgeryDuration, knots = c(550)) + SteroidUsage,
    data = core_temperature_model,
    family = binomial()) 

partial_residuals(updated_fit) |>
  mutate(.predictor_name = case_when(
    .predictor_name == "Age" ~ "Age (in Years)",
    .predictor_name == "BMI" ~ "Body Mass Index (kg/m^2)",
    .predictor_name == "SurgeryDuration" ~ "Surgery Duration (in Minutes)",
    .predictor_name == "TWATemp" ~ "TWA Core Temperature (in Celsius)",
    TRUE ~ .predictor_name
  )) |>
  ggplot((aes(x = .predictor_value, y = .partial_resid)))+
  geom_point(size = 0.3, color = 'grey50')+
  geom_line(aes(y = .predictor_effect), linewidth = 0.3)+
  geom_smooth(method = "loess", linetype = "twodash", fill = "skyblue3")+
  facet_wrap(~.predictor_name, scales = "free")+
  labs(y = "Partial Residual",
       x = "Predictor Value")+
  theme_minimal()+
  theme(strip.text = element_text(face = "bold"))
```

**Randomized Quantile Residuals:**

A third method is to use randomized quantile residuals. We can use plots of randomized quantile residuals against predictors and/or fitted values, as with any other residuals, to check the overall fit of our model. 

  - If the model is correctly specified, we would want the conditional mean of the randomized quantile
    residuals to be approximately $0.5$.

It is also useful to check that their distribution is indeed uniform (e.g. with a Q-Q plot). When the model is incorrectly specified, the distribution will not be uniform, producing patterns on the residual plots that can be interpreted. 

**Example:** Below is an example with Pima women data from the `MASS` package.


```{r}
library(MASS)
Pima.tr$pregnancy <- factor(
  ifelse(Pima.tr$npreg > 0, "Yes", "No")
)

pima_fit <- glm(type ~ pregnancy + bp, data = Pima.tr,
                family = binomial())

pima_aug <- augment_quantile(pima_fit)

p1 <- pima_aug |>
  ggplot(aes(x = .fitted, y = .quantile.resid))+
  geom_point()+
  geom_hline(yintercept = 0.5)+
  geom_smooth(fill = 'skyblue2')+
  labs(x = 'Fitted Value', y = 'Randomized Quantile Residual')+
  theme_bw()

p2 <- pima_aug |>
  ggplot(aes(sample = .quantile.resid))+
  geom_qq_line(linewidth = 0.4, distribution = stats::qunif) +
  geom_qq(size = 0.75, distribution = stats::qunif)+
  labs(x = 'Theoretical Quantiles',
       y = 'Observed Quantiles')+
  theme_bw()

p1 + p2
```

## Calibration Plots

Roughly speaking, a calibrated model is one whose predicted probabilities are accurate. For example, if the model predicts $\text{Pr}(Y = 1 \mid X = x) = 0.8$ for a particular $x$, and we observe many responses with that $x$, about $80\%$ of those responses should be $1$ and $20\%$ should be $0$.

```{r}
#| label: fig-calibration-plot
calibration_data <- data.frame(
  x = predict(pima_fit, type = "response"),
  y = ifelse(Pima.tr$type == "Yes", 1, 0)
)

ggplot(calibration_data, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Predicted probability", y = "Observed fraction") +
  ylim(0, 1)+
  theme_bw()
```

We start by converting the response from a factor to 0 or 1, so the smoother can take the average. @fig-calibration-plot shows that the calibration looks good across most of the range in our data, but as the predicted probability gets over 0.6, we see odd behavior; examining the plotted points, it appears this is due to a few observations with high predicted probability but $Y = 0$. We should not get too worried over two or three observations, so this is not particularly concerning.

*But perfect calibration does not make a good model. Calibration hence should be used together with other measures of the adequacy of the model fit, not on its own!*

## Hypothesis Tests

Wald tests are what `tidy()` defaults to for GLMs. Wald tests use a $z$ test statistic and, by default `tidy()` tests $\beta = 0$.

### Wald Tests

First, we can find a confidence interval for a probability for a specific set of regressors:

```{r}
ilogit <- function(x) 1 / (1 + exp(-x))

x <- data.frame(bp = 80,
                pregnancy = "Yes")

pred_lp <- predict(pima_fit, newdata = x,
                   type = "link", se.fit = TRUE)

ilogit(pred_lp$fit + (
  qnorm(p = c(0.025, 0.975)) *
    pred_lp$se.fit
))
```


### Confidence Intervals

We can also derive the confidence intervals for the coefficients in our model using Wald tests just using `confint()`

```{r}
exp(confint(pima_fit)) # exponentiate puts it of the odds scale
```

**Example interpretation:** A 95% confidence interval for the odds ratio associated with having a pregnancy is $[0.272, 1.47]$. Since the confidence interval overlaps with $1$, we cannot conclusively say whether prior pregnancy is associated with an increase or decrease in odds.

### Deviance Tests

Deviance tests are used for nested models and thus equivalent to an F test in linear regression. Specifically:

$$\text{Dev}_{\text{reduced}} - \text{Dev}_{\text{full}} \overset{\text{d}}{\rightarrow} \chi^2_q, \quad \text{where } q \text{ is the diff in the degrees of freedom}$$


**Example code:**

```{r}
pima_larger_fit <- glm(type ~ pregnancy + bp + age + glu,
                       data = Pima.tr, family = binomial())

anova(pima_fit, pima_larger_fit, test = "Chisq") #chisq is same as LRT
```
Here, $q = 2$. Since the p-value is significant we reject the null hypothesis that both age and glucose status have coefficients equal to one (meaning at least one is associated with a change in the odds of diabetes status), at a significance level of 0.05.

## Coefficient Interpretation

```{r}
gt(tidy(pima_fit)) |> 
  fmt_number(decimals = 3) |>
  cols_align_decimal(c(estimate, p.value))
```


**Note the above is on the log-odds scale**

```{r}
#| label: tbl-logistic-coef
#| tbl-cap: ADD

tbl_regression(pima_fit, exponentiate = TRUE) |>
  as_gt() |>
  cols_align_decimal(c(estimate, p.value))
```

**Example interpretation:** @tbl-logistic-coef gives the results of the logistic regression fit. Women with prior pregnancies were less likely to have diabetes ($OR = 0.63$, 95% CI $[0.27, 1.5]$), but this result was not statistically significant ($z = -1.1, p = 0.27$). A larger sample may be necessary to determine if a relationship exists in the population.

**Another one:** Each unit of increase in blood pressure (measured in mm Hg) is associated with an increase in the log-odds of diabetes of $0.04$, or a multiplication in the odds of diabetes by $1.04$ (95% CI $[1.01, 1.07]$).


## Outliers

We can approximate **Cook's distance** fairly well to estimate the influence of each observation on our coefficient estimates, using augment just like before. 

We could also likely **remove the outliers and see if it qualitatively changes our findings** if we aren't sure.

# Other Exponential Families

## Assumptions

When we fit a generalized linear model (such as Binomial and Poisson models), we make three key assumptions:

(1) The observations are conditionally independent given $X$

(2) The response variable follows the chosen distribution 

(3) The mean of the response is related to the predictors through the chosen link function and functional form

**Residual diagnostics can be used to check the latter two assumptions.**

## Diagnostics

We can use similar diagnostics to what we used for logistic regression. A good precursor to these is an **empirical link plot**, which would be done in the EDA section. 

**Partial Residuals:**

```{r}
ants <- read.csv("data/ants.csv")

ants_fit <- glm(Srich ~ Latitude + Elevation + Habitat, data = ants,
                family = poisson())

partial_residuals(ants_fit) |>
  ggplot(aes(x = .predictor_value, y = .partial_resid)) +
  geom_point(color = "grey40", size = 0.5) +
  geom_smooth(color = "blue", fill = 'skyblue2',
              linewidth = 0.7, linetype = "twodash") +
  geom_line(aes(y = .predictor_effect), color = 'red') +
  facet_wrap(vars(.predictor_name), scales = "free", ncol = 2) +
  labs(x = "Predictor value", y = "Partial residual")+
  theme_bw()+
  theme(strip.background = element_rect(fill = "white", color = "white"),
        strip.text = element_text(color = "black", face = "bold"),
        axis.text = element_text(color = "black"),
        plot.title = element_text(face = "bold", size = 11))
```


**Randomized Quantile Residuals:**

```{r}
library(agridat)

seed_fit_1 <- glm(cbind(germ, n - germ) ~
                    extract + gen,
                  data = crowder.seeds, family = binomial())

seed_aug <- augment_quantile(seed_fit_1)

p1 <- seed_aug |>
  ggplot(aes(x = .fitted, y = .quantile.resid))+
  geom_point()+
  geom_hline(yintercept = 0.5)+
  geom_smooth(fill = 'skyblue2')+
  labs(x = 'Fitted Value', y = 'Randomized Quantile Residual')+
  theme_bw()

p2 <- seed_aug |>
  ggplot(aes(sample = .quantile.resid))+
  geom_qq_line(linewidth = 0.4, distribution = stats::qunif) +
  geom_qq(size = 0.75, distribution = stats::qunif)+
  labs(x = 'Theoretical Quantiles',
       y = 'Observed Quantiles')+
  theme_bw()

p1 + p2

```


## Binomial Regresson

The binomial distribution is a common response distribution whenever outcomes are binary, or whenever we count a certain binary outcome out of a total number of trials.

The binomial distribution is suitable when there is a fixed and known total number of trials. Each observation hence consists of a number of successes and a total number of trials; the number of trials may differ between observations. 

The model is (with a logit-link function, the default):

$$n_iY_i \mid X_i = x_i \sim \text{Binomial}(n_i, \text{logit}(\beta^\intercal x_i))$$
Hence, the sample proportion $Y_i$ should be proportional to $\text{logit}(\beta^\intercal x_i)$. 

   - **$Y_i$ is the rate/probability of success!**

In `R`, a binomial response variable for $n > 1$ can be provided as a a two-column matrix with the number of successes and the number of failures.

```{r}
#| echo: true
#| code-fold: false

library(agridat)

seed_fit_1 <- glm(cbind(germ, n - germ) ~
                    extract + gen,
                  data = crowder.seeds, family = binomial())
seed_fit_2 <- glm(cbind(germ, n - germ) ~
                    extract * gen,
                  data = crowder.seeds, family = binomial())
```


### Inference and Predictions

**Crowder Seed Example:**

```{r}
crowder.seeds |>
  mutate(germ_rate = germ/n) |>
  ggplot(aes(x = gen, y = germ_rate, color = extract))+
  geom_boxplot()+
  labs(y = 'Germination Rate', x = 'Seed Variety')+
  theme_bw()
```


```{r}
seed_test <- anova(seed_fit_1, seed_fit_2, test = "Chisq")
seed_test
```

Based on the analysis of deviance table, we conclude that that extract differs by side type, $\chi^2(1) = 6.41, p = 0.011$. 

  - Like with logistic regression, we can use `tbl_regression()` to compute the coefficients in each model. *E.g., for the interaction model:*

```{r}
#| echo: true
#| code-fold: false
tbl_regression(seed_fit_2, exponentiate = TRUE)
```

  -   We can also compute the predictions and the standard errors. From the table, we see that the '75 seeds with cucumber extract have the highest predicted probability of germination.

```{r}
#| echo: true
#| code-fold: false

# first we create a grid of all possible versions of the relevant regressors we want to get predictions for:
xs <- expand.grid(gen = c("O75", "O73"),
                  extract = c("bean", "cucumber"))

predictions <- predict(seed_fit_2, newdata = xs,
                       type = "response", se.fit = TRUE) # response = rate scale

gt(data.frame(gen = xs$gen, extract = xs$extract,
           pred_germ_rate = predictions$fit, se = predictions$se.fit)) |>
  fmt_number(decimals = 3)
```


#### Interpretation

```{r}
library(Sleuth3)

island_fit <- glm(cbind(Extinct, AtRisk - Extinct) ~ log10(Area), data = case2101,
    family = binomial())

exp(coef(island_fit))

ci <- exp(confint(island_fit)['log10(Area)',])
ci
```

**Interpretation:** a ten-fold increase in island size (i.e., an island that is 10 times larger) is associated with the odds of extinct being multiplied by 0.5 (95% CI [$0.39, 0.64$]) *(if not on log scale, would be a one-unit increase in $X$)*.

## Poisson Regression

If a certain event occurs with a *fixed rate*, and the events are *independent* (so that the occurrence of one event does not make another more or less likely), then the count of events over a fixed period of time will be Poisson-distributed. This makes Poisson GLMs well-suited for *response variables that are \underline{counts}*.

The canonical link for a Poisson distribution is the **log-link**, meaning that under the Poisson model, we have:

$$\log(\mathbb{E}[Y \mid X = x]) = \beta^\intercal x$$
which is equivalent to:

$$\mathbb{E}[Y \mid X = x] = \exp(\beta^\intercal x)$$

**Thus, the variance of $Y$ depends on $X$ in Poisson GLMs!**


### Inference and Predictions

**Ant Example:**

```{r}
# omitted EDA but idea is to see which covariates are associated with the ant species counts in the data

ants_fit <- glm(Srich ~ Latitude + Elevation + Habitat, data = ants,
                family = poisson())

tbl_regression(ants_fit, exponentiate = TRUE)
```


**Example Interpretation:** One additional degree in latitude is associated with the mean number of ant species being *multiplied* by 0.79, holding all other covariates in the model constant.

We can also find confidence intervals by exponentiating the confidence interval based on the profile likelihood:

```{r}
#| echo: false
#| code-fold: false

exp(confint(ants_fit))
```


**Another example (Howard the Duck):** Increasing the fraction of time a duck spends at the park by $0.1$ is associated with the mean number of hospitalizations being multiplied by $1.09$ (95% CI [$1.06, 1.13$]), holding all other covariates in the model constant.

### Offsets

Offsets are useful when the observed counts recorded for different population sizes or time periods. 

**Smoking Example:**

```{r}
smokers <- read.csv("data/smokers.csv")

smokers |>
  group_by(age, smoke) |>
  summarize(Deaths = sum(deaths),
            `Person-years` = sum(py)) |>
  ungroup() |>
  gt()
```

The person-years column indicates that people in the smoking study were observed for different number of years and/or there were different numbers of people in the study who were in each age range. 

If we are interested in death *rate*, we could model the rate directly as some function of our covariates. 

```{r}
smokers |>
  mutate(death_rate = deaths/py) |>
  ggplot(aes(x = age, y = death_rate))+
  geom_point(aes(color = smoke, shape = smoke))+
  theme_bw()
```

In turn, we could model the outcomes (deaths in our case) as approximately Poisson, given the death rate $f(\beta^\intercal X)$:

$$Y \sim \text{Poisson}(\exp(\beta^\intercal X) + \log(P))$$
where $P$ is the offset, i.e., a term in our model that is fixed to have coefficient one rather than a slope that is estimated.

**Note for deaths, we are technically approximately Poisson as the upper-bound on deaths is the number of people in our study. That said, so long as the death rate is low, the Poisson distribution will assign very little probability on impossible death numbers**

If the model with the offset is true, then we would expect a linear relationship between $\log(Y/P)$ and $X$. We can use the `offset` argument in the `glm()` function to specify our offset:

```{r}
#| echo: true
#| code-fold: false

smoke_fit <- glm(deaths ~ (age + I(age^2)) * smoke,
                 offset = log(py), #offset!
                 data = smokers, family = poisson(link = "log"))
```

Further, we can use deviance tests with offsets 

```{r}
#| echo: false
#| code-fold: false
additive_smoke <- glm(deaths ~ age + smoke, offset = log(py),
                      data = smokers, family = poisson(link = "log"))

anova(additive_smoke, smoke_fit, test = "Chisq")
```

**Risk Ratio Calculation:**

Now suppose we want to compare smokers and non-smokers at age 40, calculating a *risk ratio*: the ratio of risk of death for smokers versus non-smokers. A simple way would be to make a prediction and take the ratio. To predict the rate of death, we can predict the mean number of deaths for one person-year:

```{r}
new_smokers <- data.frame(smoke = c("no", "yes"),
                          age = c(40, 40),
                          py = c(1, 1))
preds <- predict(smoke_fit, newdata = new_smokers, type = "response")
preds[2] / preds[1]

# or with:
coef_vec <- c(0, 0, 0, 1, 40, 1600)

exp(sum(coef_vec * coef(smoke_fit)))
```


The standard errors are not independent. We can create the Wald confidence interval by leveraging the fact that the variance and mean are equal in Poisson models.

```{r}
se <- sqrt(t(coef_vec) %*% vcov(smoke_fit) %*% coef_vec)[1,1]

bounds <- sum(coef_vec * coef(smoke_fit)) + c(-2, 2) * se
exp(bounds)
```

The quite a large confidence interval is indicative of the difficulty in estimating risk ratios when the underlying rate of events is so low.

## Overdispersion

Overdispersion is when there is more variance in $Y$ than the response distribution would predict. This could be due to:

- Insufficient predictors. That is, there might be other factors associated with the expected value of $Y$ that we do not observe

- There might be correlations we did not account for (e.g., a binomial distribution assumes the $n$ trials are independent but what if success in one is correlated with increased success in the others?)

A remedy for overdispersion is to use quasi-likleihood. These can be fit using `quasibinomial()` and `quasipoisson()` families in the `glm()` function. Compared to non-quasi models, the estimates of the coefficients will be the same, **only the confidence intervals differ, witht heir width expanding by a constant factor**

**We can check for overdispersion using Q-Q plot of randomized quantile residuals**

**Example:** returning to the seed data:

```{r}
#| echo: false
#| code-fold: false

seed_quasi <- glm(cbind(germ, n - germ) ~
                    extract * gen,
                  data = crowder.seeds,
                  family = quasibinomial())

tbl_regression(seed_quasi, exponentiate = TRUE)
```


**Important Note:** since overdispersed models no longer specify a distribution of the outcome conditional on the covariates, tests based on the log-likelihood are no longer valid and we can no longer use randomized quantile residuals or do line-ups. 

**We can still still do Wald Tests and interpret coefficients using the new Confidence Interval**

*Estimating the dispersion parameter...*

```{r}
#| echo: true
#| code-fold: false
sum(residuals(ants_fit, type = 'deviance')^2)/ants_fit$df.residual
```


## Outliers

We can approximate **Cook's distance** fairly well to estimate the influence of each observation on our coefficient estimates, using augment just like before. 

We could also likely **remove the outliers and see if it qualitatively changes our findings** if we aren't sure.

```{r}
augment(ants_fit) |>
  ggplot(aes(x = .fitted, y = .cooksd))+
  geom_point()+
  theme_bw()
```


# Bootstrapping (Brief Aside)

So far, when we have discussed inference (such as hypothesis tests and confidence intervals), we have used it to try to learn about the underlying population relationships, such as to do inference on the factors associated with some outcome. 

To do this inference, we relied on mathematical derivations of the distribution of estimators and statistics. These derivations rely on certain \underline{assumptions} about the data and the population it is drawn from, and if those assumptions are not true, the derivations are not valid. 

Boostrapping is an alternative way to esitmate the distribution of estimators and statistics. This can lead to more accurate uncertainty quantification for coefficients (that said, if the sample is not iid, then resampling from the observed data is not likel obtaining a sample from the population...)

# Penalized Models

  - Helpful if we care about prediction over inference and thus want to minimize the total error (which is a combination of bias and variance).

-   Penalization methods are also helpful when we have $p > n$ since that have a unique solution even with a higher dimension of regressors than observations.

## Ridge Regression

Ridge Regression penalizes the L-2 norm of the covariates. We can implement Ridge Regression using the `glmnet` package, which requires us to provide $X$ as a matrix and $Y$ as a vector of responses. Note that `glmnet()` standardizes each column to have variance 1 automatically.

*[For Ridge Regression, $\alpha = 0$]*

```{r}
library(mvtnorm)

sparse_pop <- population(
  x = predictor(rmvnorm,
                mean = rep(0, 100),
                sigma = diag(100)),
  y = response(4 + 0.2 * x1 - 5 * x2 + x3 + 0.4 * x4 + 8 * x5,
               error_scale = 1)
)

sparse_samp <- sparse_pop |>
  sample_x(n = 90) |>
  sample_y()

x <- model.matrix(~ . - 1 - y, data = sparse_samp)

ridge_fit <- glmnet(x, sparse_samp$y, alpha = 0)

plot(ridge_fit)
```

To determine which penalty parameter to select, we can use **cross-validation** via `cv_glmnet()` (which by default does 10-fold CV).

```{r}
cv_results <- cv.glmnet(x, sparse_samp$y, alpha = 0)
cv_results$lambda.min

# if we want to plot to see what is happening:
plot(cv_results)
```

While the coefficients of Ridge Regression are not fully **interpretable**, we can look at which predictors have the largest magnitudes to get a sense of the relative effect size between different covariates and our outcome variable.

```{r}
coefs <- coef(ridge_fit, s = cv_results$lambda.min)
coefs[order(abs(coefs[, 1]), decreasing = TRUE)[1:5], ]
```

**Ridge Regression is particularly helpful when there are collinear predictors -- which result in high prediction variance** (encourages the effects to be ``shared" between collinear predictors). So, Ridge Regression **reduces the variance of predictions**.

## Lasso

Lasso is another penalized regression model. Its most useful propoerty is that (depending on the value of the penalization parameter), it forces many of the coefficient estimates to be exactly zero. That is, **Lasso promotes sparsity**, which matches real-world populations where we expect the true coefficients to be sparse (*think: genetics*)

We can implement Lasso with `glmnet` (*the only difference is now we set $\alpha = 1$*). 

```{r}
lasso_fit <- glmnet(x, sparse_samp$y, alpha = 1)

plot(lasso_fit)
```

Like before, we can also cross-validate:

```{r}
cv_results <- cv.glmnet(x, sparse_samp$y, alpha = 1)
cv_results$lambda.min

# and plot if we want a bitter idea of what is happening
plot(cv_results)
```

and, we also can look at the coefficients with the largest magnitude like before:

```{r}
coefs <- coef(lasso_fit, s = cv_results$lambda.min)
coefs[order(abs(coefs[, 1]), decreasing = TRUE)[1:15], ]
```

**Importantly**, it can be shown that if we know the right penalization value, the model selection by Lasso is model selection consistent (if chooses all and only the true non-zero coefficient with probability 1 as the sample size tends to infinity).

While we choose our penalization value based on cross-validation, so model selection consistency might not hold.

## The Elastic Net

Elastic net generalizes the case of Ridge Regression and Lasso by allowing both $\alpha$ and $\lambda$ to be hyperparameters, the combination of which trades off the benefits of Ridge Regression and Lasso. **That is, with a $\alpha$ closer to one, more sparsity is induced, but with a $\alpha$ closer to 0, variance from collinearity is reduced more.**

*To cross-validate over both lambda and alpha, you have to manually loop over alpha values yourself, and run cv.glmnet() inside that loop.*

## Lasso Model Example (on Genetic Data)

```{r}
#| echo: true
#| code-fold: false

set.seed(47)
genedat <- read_csv('data/genedat-exam.csv')

x <- model.matrix(Disease ~ . -1 -CaseId -X, data = genedat) 
Y <- if_else(genedat$Disease == 'control', 0, 1)

cv_genedat <- cv.glmnet(x, Y, alpha = 1, family = "binomial") 
#ideally would have done group cv here

fit_genedat <- glmnet(x, Y, family = "binomial", alpha = 1)

coefs <- coef(fit_genedat, s = cv_genedat$lambda.min)

results_df <- data.frame(coeff_est = coefs[,1]) |>
  rownames_to_column('covariate') |>
  filter(coeff_est != 0, !grepl('Intercept', covariate)) |>
  arrange(desc(abs(coeff_est))) |>
  slice_head(n = 10) |> 
  gt() |>
  fmt_number(decimals = 2) |>
  cols_align_decimal(coeff_est)
  
# ignoring within person correlation for now, but would likely want to  subset data to just one cortex area especially since there isn't two observations (could also do mixed effects approach as a future step). Also, if goal is prediction, we would want to have test and train split sets -- probably by group to avoid data leakage
```


```{r}
#| echo: true
#| code-fold: false
set.seed(123)
folds <- group_vfold_cv(genedat, group = CaseId, v = 5)
lasso_errors <- sapply(folds$splits, function(split) {
  train <- analysis(split)
  train_x <- model.matrix(~ . -1 -X -Disease -CaseId, data = train)
  
  test <- assessment(split)
  test_x <- model.matrix( ~ . -1 -X -Disease -CaseId, data = test)
  
  fit_genedat <- glmnet(train_x, train$Disease, family = "binomial", alpha = 1)
  
  pred <- predict(fit_genedat, newx = test_x, s = cv_genedat$lambda.min,
                   type = 'response')
  
  pred_class <- if_else(pred >= 0.5, 'autism', 'control')
  
  mean(pred_class == test$Disease)
})

mean(lasso_errors)
```


# Prediction

## Splitting the Data

### General Data Splitting

```{r}
#| echo: true
#| code-fold: false
set.seed(47)
car_split <- initial_split(mtcars) #default is 3:1 training-test split
train_data <- training(car_split)
test_data <- testing(car_split)
```

### With Stratification

Including a strata is helpful, particularly if there is class imbalance in our covariates or outcome to the point where it is possible that training and/or test won't have the same levels of the variables

```{r}
#| echo: true
#| code-fold: false
car_split <- initial_split(mtcars, strata = 'cyl')
train_data <- training(car_split)
test_data <- testing(car_split)
```

### By Time

```{r}
#| echo: true
#| code-fold: false
covid_hew <- read_csv('data/covidcast-hew.csv')

training <- covid_hew |>
  mutate(year_value = year(time_value),
         month_value = month(time_value)) |>
  filter(month_value == 1)

testing <- covid_hew |>
  mutate(year_value = year(time_value),
         month_value = month(time_value)) |>
  filter(month_value == 3)
```

### By Group

Helpful if there is correlation between observations which would lead to data leakage concerns if we do not split the data by group. 

```{r}
#| echo: true
#| code-fold: false
car_split <- group_initial_split(mtcars, group = cyl)
train_data <- training(car_split)
test_data <- testing(car_split)
```

## Nested Cross Validation

Often, we might need to do cross-validation and want to keep a hold out set that we can test our model on later. Here would be a a framework for nested cross validation:

(1) Split the full dataset into testing and training (we might do this randomly, by time, by group, etc.)

(3) On the training data do k-fold cross validation to select any needed parameters (e.g., penalization parameters)

(4) Fit the model on the full training data with the selected tuning parameter values

(5) Assess predictive performance on the test data 


## Useful Metrics

*Report on test data but could be interesting to compare test performance to how it performs on training data*

- If test data performance is same or better than training data, this indicates that the model is underfit/biased

- If the test data performance is much lower than the training data, this might raise concerns about overfitting

**For regression:**

- Mean-Squared Error (MSE)

- Root Mean-Squared Error (RMSE): a measure of the average magnitude of the errors between predicted and actual values in a regression model

- Actual versus Predicted Outcome Plot (ideal would be along the main diagonal)

**For classification:**

- Calibration plot

- Confusion matrix

    - False Negative Rate, False Positive Rate, Specificity, Sensitivity

- ROC Curve (AUC)

- Accuracy

*Note:* when looking at accuracy, always report what the base incidence rate is (this would be the accuracy of a null model).

# Missing Data

Steps to handling missingness:

1. Acknowledge the missingness

2. Discuss possible reasons for the missingness

3. Note the implication to the analysis

**When complete-cases analysis is not concerning:**

- When data is missing completely at random

- When missingness only depends on observed covariates

**See examples in Chapter 22**

*Some general notes:*

  - Impute if clear value that should be filled in.
  
  - Can impute to mean if needed (if there is not much variability in the covariate
  
  - Generally, we weary of imputing the response variable only it is very clear

# Multiple Testing

One concern if we are conducting several hypothesis tests is multiple testing. While not in the scope for the DA exam to correct for (likely), it could be an important limitation to mention in the report.

# Causal Claims

If we are confident in our causal model and can control for the necessary confounders, we can estimate a chosen causal path.  

**In this situation, our obligation is to make clear the limitations of our claims:**

1. If our regression model is misspecified or otherwise incorrect, our estimates may be wrong.

2. If our causal model is missing important confounders, or we have measured some confounders incompletely or inaccurately, our estimates may include some bias from confounding.

3. If our data comes from a specific sample or subset of a population, the causal claims may not generalize beyond it.

*Note: choosing confounders to allow us to estimate one coefficient causally does not necessarily imply that all the other coefficients in the model can be interpreted causally.*